\MakeUppercase{Layered $H$-Partitions of $k$-Planar Graphs} \thanks{This work was partly funded by NSERC and MRI.}

### 2019-05-17 k-planar.tex
Fixing an off-by-one error
Fixed off-by-one error added example for k=1

@@

We prove the following result:
\begin{thm}\thmlabel{k-planar}
  Every $k$-planar graph has a layered $H$-partition of layered-width at most $24k^2 + 36k + 12$ in which $H$ has treewidth at most $k^3/6+k^2/2+k/3-1$.
\end{thm}

@@

\begin{thm}\thmlabel{g-k-planar}
  Every $(g,k)$-planar graph has a layered $H$-partition of layered-width at most $\max\{16gk^2 + 24gk + 8g, 24k^2 + 36k + 12\}$ in which $H$ has treewidth at most $(3^{k+2}-3)/2$.

@@

The remainder of the paper is organized as follows: In \secref{k-planar} we prove \thmref{k-planar}.  In \secref{1-planar} we prove \thmref{1-planar}.  
In \secref{genus-g} we prove \thmref{g-k-planar}.

@@

\begin{clm}\clmlabel{bag-size}
  The tree decomposition $\{B_x: x\in V(K)\}$ of $H$ contains no bag larger than $k^3/6 + k^2/2 + k/3$.
\end{clm}

@@

  Refer to \figref{d0d1}.  We will now prove that 
  \begin{equation}
     |\{\delta\in\{d_1,\ldots,d_0-1\}:  \mbox{there exists a $\delta$-walk} \}| \le \binom{k+1}{2} \enspace . \eqlabel{one-round}
  \end{equation}
  \begin{figure}
    \begin{center}
      \includegraphics{figs/d0d1}
    \end{center}
    \caption{$F_{d_1},\ldots,F_{d_0}$ and the resulting graph $I$.  For readability, the edges $\{id_1:i\in\{d_1-2,\ldots,d_0\}$ are not shown.}
    \figlabel{d0d1}
  \end{figure}

@@

  Therefore, the total number of $\delta\in\{d_1+1,\ldots,d_0\}$ such that there exists a $\delta$-walk is at most
  \[
     \sum_{t=0}^{k+1} (k+2-t) = \sum_{t=1}^{k+2} t = \binom{k+1}{2}
  \]
  We are now very close to \eqref{one-round}, which asks for the number of $\delta\in\{d_1,\ldots,d_0-1\}$ such that there exists a $\delta$-walk.
  The preceding formula counts $\delta=d_0$, but does not count $\delta=d_1$.  These two effects cancel each other, thus establishing \eqref{one-round}.

@@

    \item $\ell\le k+1$, so $\ell-i \le \ell-s \le k+1-s$.
  \end{compactenum}
  Exactly the same analysis used to establish \eqref{one-round} shows that
  \begin{equation}
     |\{\delta\in\{d_{s+1},\ldots,d_s-1\}:  \mbox{there exists a $\delta$-walk} \}| \le \binom{k+1-s}{2} \enspace . \eqlabel{one-round-s}
  \end{equation}

  We conclude that the maximum number of $\delta\in\{0,\ldots,d-1\}$ such that there exists a $\delta$-walk is at most
  \[
      \sum_{s=0}^{k+1} \binom{k+1-s}{2}
      = \sum_{s=0}^{k+1} \binom{s}{2}
      = \frac{k^3}{6} + \frac{k^2}{2} + \frac{k}{3} \ge |B_x| \enspace . \qedhere
  \]
\end{proof}

@@

  By \clmref{bag-size}, $\mathcal{S}$ produces a graph $H=G/\mathcal{S}$ of treewidth at most $k^3/6+k^2/2+k/3-1$.  By \clmref{width-of-g-plus}, for each $x\in V(H)$ and each integer $i$, $|B_x\cap V_i|\le 24k+12$, so $|B_x\cap V_i'|\le (k+1)(24k + 12) = 24k^2+36k+12$. 
\end{proof}

>>>>

\newcommand{\Bagsize}{\ensuremath{\frac{k^3}{6} + \frac{3k^2}{2} + \frac{13k}{3} + 4}}
\newcommand{\bagsize}{\ensuremath{k^3/6 + 3k^2/2 + 13k/3 + 4}}
\newcommand{\treewidth}{\ensuremath{k^3/6 + 3k^2/2 + 13k/3 + 3}}

@@

We prove the following result:
\begin{thm}\thmlabel{k-planar}
  Every $k$-planar graph has a layered $H$-partition of layered-width at most $24k^2 + 36k + 12$ in which $H$ has treewidth at most $\treewidth$.
\end{thm}

@@

\begin{thm}\thmlabel{g-k-planar}
  Every $(g,k)$-planar graph has a layered $H$-partition of layered-width at most $\max\{16gk^2 + 24gk + 8g, 24k^2 + 36k + 12\}$ in which $H$ has treewidth at most $\treewidth$.

@@

The remainder of the paper is organized as follows: In \secref{k-planar} we prove \thmref{k-planar}.  In \secref{1-planar} we prove \thmref{1-planar}.  
In \secref{g-k-planar} we prove \thmref{g-k-planar}.

@@

\begin{clm}\clmlabel{bag-size}
  The tree decomposition $\{B_x: x\in V(K)\}$ of $H$ contains no bag larger than $\bagsize$.
\end{clm}

@@

  Refer to \figref{d0d1}.  We will now prove that 
  \begin{equation}
     |\{\delta\in\{d_1+1,\ldots,d_0\}:  \mbox{there exists a $\delta$-walk} \}| \le \binom{k+3}{2} \enspace . \eqlabel{one-round}
  \end{equation}
  \begin{figure}
    \begin{center}
      \includegraphics{figs/d0d1}
    \end{center}
    \caption{$F_{d_1},\ldots,F_{d_0}$ and the resulting graph $I$.} 
    \figlabel{d0d1}
  \end{figure}

@@

  Therefore, the total number of $\delta\in\{d_1+1,\ldots,d_0\}$ such that there exists a $\delta$-walk is at most
  \[
     \sum_{t=0}^{k+1} (k+2-t) = \sum_{t=1}^{k+2} t = \binom{k+3}{2}
  \]
  This establishes \eqref{one-round}.  An illustration of this analysis, showing the case $k=1$ is shown in \figref{one-round-example}.
  \begin{figure}
    \begin{center}
      \includegraphics{figs/one-round-example}
    \end{center}
    \caption{The set of $6=\binom{k+3}{2}$ vertices in $I^+$ reachable by paths of walks of length at most $k+1$ (for $k=1$).}
    \figlabel{one-round-example}
  \end{figure}

@@

    \item $\ell\le k+1$, so $\ell-i \le \ell-s \le k+1-s$.
  \end{compactenum}
  Exactly the same analysis used to establish \eqref{one-round} shows that
  \begin{equation}
     |\{\delta\in\{d_{s+1}+1,\ldots,d_s\}:  \mbox{there exists a $\delta$-walk} \}| \le \binom{k+3-s}{2} \enspace . \eqlabel{one-round-s}
  \end{equation}

  We conclude the proof with 
  \begin{align*}
    |B_x| & \le |\{\delta\in\{0,\ldots,d\}:  \mbox{$\exists$ a $\delta$-walk} \}| \\
      & = \sum_{s=0}^{k+1}|\{\delta\in\{d_{s+1}+1,\ldots,d_s\}:  \mbox{$\exists$ a $\delta$-walk} \}| \\
      & \le\sum_{s=0}^{k+1} \binom{k+3-s}{2} \\
      & = \sum_{s=2}^{k+3} \binom{s}{2} \\
      & = \frac{1}{6}(k+2)(k+3)(k+4) \\
      & = \frac{k^3}{6} + \frac{3k^2}{2} + \frac{13k}{3} + 4  \enspace . \qedhere
  \end{align*}
\end{proof}

We now have all the pieces needed to complete the proof of \thmref{k-planar}.

@@

  By \clmref{bag-size}, $\mathcal{S}$ produces a graph $H=G/\mathcal{S}$ of treewidth at most $\treewidth$.  By \clmref{width-of-g-plus}, for each $x\in V(H)$ and each integer $i$, $|B_x\cap V_i|\le 24k+12$, so $|B_x\cap V_i'|\le (k+1)(24k + 12) = 24k^2+36k+12$. 
\end{proof}
